Leveraging The Progress in Automated Theorem Proving for Program Synthesis\\ \small{WWTF ICT 2022}

### 2022-03-27 WWTF/Draft Alexander/Proposal.tex
Formatting and packages in Proposal.tex
expected results, novelty and relevance

\documentclass[onehalfspacing]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{ wasysym }
\usepackage{ stmaryrd }
\usepackage{ mathpartir }
\usepackage{xcolor}
\usepackage{bussproofs}
\usepackage[left=3.00cm, right=3.00cm, top=3.00cm,bottom=3.0cm]{geometry}
\usepackage{pdflscape}
\usepackage{xspace}

@@

	As outlined we expect to obtain the necessary theoretical foundations as well as a practical system to leverage state-of-the-art classical theorem provers for program extraction. At the current time no such system exists. Much of the theoretical groundwork for implementing such a system is already present, but is scattered and mostly applied in  theoretical model settings.

	A successful completion of this project would lay the foundation for verified program synthesis via automated theorem provers.

>>>>

\documentclass[a4paper,12pt]{article}

@@

	We now give details on (WP1). As mentioned in the previous section there are currently two main approaches in the literature for obtaining programs from classical proofs: 1) Parigot's $\lambda\mu$-calculus~\cite{Parigot1} gives a direct translation of classical proofs with datatypes and is essentially ready for implementation. However, since non-intuitionistic tautologies like $\neg\forall xP(x)\to \exists x\neg P(x)$ or $A\vee \neg A$ are assigned a program, we lose the realizability of $\exists$ and $\vee$. As discussed in~\cite{practical}, in practice it is often possible to eliminate the non-intuitionistic fragments. We aim to establish stronger theoretical guarantees on when this is possible but also more practical benchmarks on the feasibility of this approach, even when realizability is not ensured by a theorem. 2) Sometimes classical validity guarantees intuitionistic validity, most famously for $\Pi_2$-formulas in Arithmetic~\cite{Friedman}. There have been various generalizations of this result, most notably by Schwichtenberg's group~\cite{schwichtenberg}. We will adapt their work to our more general setting.

	We expect that in practice these approaches will be sufficient in many situations. However we also suggest a completely new direction: We aim for a translation between provers rather than proofs, so we examine what can be learned by attempting proofs of modified versions of the target formula. A hopeful sign is the reverse direction: given an intuitionistic prover we can establish classical provability via double negation translation. A first approximation has been outlined in~\cite{RDNT}.

	Using this theoretical groundwork we will implement a synthesis system on top of Vampire for (WP2) targeting a suitable functional programming language. We will investigate how syntactic context and redundant
	axioms can be used to restrict and guide proof search and we will examine the merit of our approach when only input/output pairs are given.

### 2022-03-27 WWTF/main.tex
expected results, novelty and relevance

\usepackage{url}

@@

\section{Introduction, background and state of the art (max. 2500 characters)}
%What is the topic and broad background of this research project? What is the state of the art in this field?
Programs are usually first written and then tested/verified.
In contrast, program synthesis allows a system developer to derive a program in one step.
At the same time, program synthesis holds the promise of freeing the programmer from low-level details and of enabling the programmer to focus on the high-level intent.
Program synthesis is a long-standing problem that crosses many areas of computer science and mathematics including programming languages, AI, formal methods and proof theory.
Traditionally, program synthesis has been formalized as a problem in deductive theorem proving~\cite{conf/ijcai/MannaW79}:
Given a logical formula that states that for all inputs there exists an output such that the required specification holds, one searches for a constructive proof of this formula and then extracts a program from this proof.
A classic route to program extraction is the \emph{Curry-Howard correspondence}~\cite{?} (also known as \emph{proofs-as-programs}) where one exploits the equivalence between proofs in intuitionist logic and functional programs in order to directly extract a functional program from an intuitionistic proof.
The Curry-Howard correspondence denotes a fundamental relationship between two seemingly unrelated formalisms, namely proofs systems on one hand and computational models on the other hand.
The Curry-Howard correspondence has been a major driving force for research spanning functional programming languages, type theory, proof systems and constructive logic.
However, progress on the deductive synthesis problem has been slow for many years because of a lack of automated and scalable provers.
Instead, a recent variant of the program synthesis problem has enabled dramatic progress (we refer to the special issue~\cite{fisman2022special} and the workshop\footnote{\url{https://simons.berkeley.edu/workshops/tfcs2021-1}} for an overview):
%\footnote{https://www.cs.technion.ac.il/~shaull/SYNT2022/index.html}
The logical specification is accompanied by a \emph{syntactic template} that constrains the space of possible implementations.
The synthesis procedure then searches for/enumerates candidate implementations and either rejects these candidates by finding violating input/output pairs or accept the generated program by verifying their correctness.
The syntactic approach is very versatile and synthesis approaches have been explored that solely start from input/output examples or that mine possible program fragments from public code repositories.
While the reported progress is impressive, we believe that the traditional deductive approach to synthesis has been unjustly neglected.

@@

The main goal of this project is to connect the classical approach to deductive synthesis --  proofs-as-programs~\cite{?}--, with modern automated theorem provers, in particular first-order solvers (and SMT solvers???), which have made impressive progress over the recent years (as evidenced by yearly competitions\footnote{\url{http://www.tptp.org/CASC/}}).

We believe this approach offers the following advantages:
1) First-order solvers are particularly suitable to solving \emph{quantified} formulas as needed for synthesis.
2) Proofs-as-programs represents a longstanding approach to program extraction.
That is to say, the groundwork for program extraction from proofs has been laid, we just need to adapt this theory to the specifics of modern automated theorem provers.
3) We are able to take advantage of the highly sophisticated and optimized proof search strategies of first-order solvers instead of needing to define and implement \emph{ad-hoc} search strategies.
%Moreover, we will profit from any future advancements in automated theorem proving.

@@

Much of the theoretical groundwork for implementing such a system is already present,
but is scattered and mostly limited to theoretical settings.
We sketch here the two main directions that connect the Curry-Howard correspondence to classical logic:
1) The Curry-Howard correspondence can be extended to classical logic using control operators~\cite{Control1}, Parigot's $\lambda\mu$-Calculus~\cite{Parigot1} or CPS-translation~\cite{CPS1}.
However in this process we must in general give up on some desirable properties, at the very least the realizability of $\vee$ and $\exists$.
2) Under certain conditions, classically valid proofs can be transformed into intuitionistic proofs, e.g. proofs of $\Pi_2$ sentences in Peano Arithmetic~\cite{HAPA}, via Friedman's Translation~\cite{Friedman} or similar methods.
We will adapt and specialise these techniques from the literature to the proofs produced by automated theorem provers such as Vampire:
A) We will develop the first\emph{ program extraction calculi} for \emph{resolution resp. superposition proofs}.
In particular, we need to pay careful attentiont to the normal-form transformation --- a pre-processing step --- in automated theorem provers, which also involve equivalences that are only classically valid.
C) We will identify assumptions under which our adaptions of the techniques 1) and 2) guarantee that such a program extraction is possible.
We are not aware of related results in the literature.

>>>>

\usepackage{url}
\usepackage{xcolor}

@@

\section{Introduction, background and state of the art (max. 2500 characters)}
%What is the topic and broad background of this research project? What is the state of the art in this field?
Programs are usually first written and then tested/verified.
In contrast, program synthesis allows a system developer to derive a program in one step.
At the same time, program synthesis holds the promise of freeing the programmer from low-level details and of enabling the programmer to focus on the high-level intent.
Program synthesis is a long-standing problem that crosses many areas of computer science and mathematics including programming languages, AI, formal methods and proof theory.
Traditionally, program synthesis has been formalized as a problem in deductive theorem proving~\cite{conf/ijcai/MannaW79}:
Given a logical formula that states that for all inputs there exists an output such that the required specification holds, one searches for a constructive proof of this formula and then extracts a program from this proof.
A classic route to program extraction is the \emph{Curry-Howard correspondence}~\cite{?} {\color{red}Standardreferenz Proofs and Types~\cite{girard1989proofs} oder Original Curry~\cite{curry1934functionality}} (also known as \emph{proofs-as-programs}) where one exploits the equivalence between proofs in intuitionist logic and functional programs in order to directly extract a functional program from an intuitionistic proof.
The Curry-Howard correspondence denotes a fundamental relationship between two seemingly unrelated formalisms, namely proofs systems on one hand and computational models on the other hand.
The Curry-Howard correspondence has been a major driving force for research spanning functional programming languages, type theory, proof systems and constructive logic.
However, progress on the deductive synthesis problem has been slow for many years because of a lack of automated and scalable provers.
Instead, a recent variant of the program synthesis problem has enabled dramatic progress (we refer to the special issue~\cite{fisman2022special} and the workshop\footnote{\url{https://simons.berkeley.edu/workshops/tfcs2021-1}} for an overview):
%\footnote{https://www.cs.technion.ac.il/~shaull/SYNT2022/index.html}
The logical specification is accompanied by a \emph{syntactic template} that constrains the space of possible implementations.
The synthesis procedure then searches for/enumerates candidate implementations and either rejects these candidates by finding violating input/output pairs or accept the generated program by verifying their correctness.
The syntactic approach is very versatile and synthesis approaches have been explored that solely start from input/output examples or that mine possible program fragments from public code repositories.
While the reported progress is impressive, we believe that the traditional deductive approach to synthesis has been unjustly neglected.

@@

The main goal of this project is to connect the classical approach to deductive synthesis --  proofs-as-programs~\cite{?}{ \color{red}s.o.}--, with modern automated theorem provers, in particular first-order solvers (and SMT solvers???), which have made impressive progress over the recent years (as evidenced by yearly competitions\footnote{\url{http://www.tptp.org/CASC/}}).

We believe this approach offers the following advantages:
1) First-order solvers are particularly suitable to solving \emph{quantified} formulas as needed for synthesis.
2) Proofs-as-programs represents a longstanding approach to program extraction.
That is to say, the groundwork for program extraction from proofs has been laid, we just need to adapt this theory to the specifics of modern automated theorem provers. {\color{red}in the classical case not everything has been said, perhaps "much groundwork..." and "...adapt and extend..."}
3) We are able to take advantage of the highly sophisticated and optimized proof search strategies of first-order solvers instead of needing to define and implement \emph{ad-hoc} search strategies.
%Moreover, we will profit from any future advancements in automated theorem proving.

@@

Much of the theoretical groundwork for implementing such a system is already present,
but is scattered and mostly limited to theoretical settings.
We sketch here the two main directions that connect the Curry-Howard correspondence to classical logic:
1) The Curry-Howard correspondence can be extended to classical logic using control operators~\cite{Control1}, Parigot's $\lambda\mu$-Calculus~\cite{Parigot1} or CPS-translation.
However in this process we must in general give up on some desirable properties, at the very least the realizability of $\vee$ and $\exists$.
2) Under certain conditions, classically valid proofs can be transformed into intuitionistic proofs, e.g. proofs of $\Pi_2$-sentences in Peano Arithmetic~\cite{HAPA}.
We will adapt and specialise these techniques from the literature to the proofs produced by automated theorem provers such as Vampire:
A) We will develop the first\emph{ program extraction calculi} for \emph{resolution resp. superposition proofs}.
In particular, we need to pay careful attentiont to the normal-form transformation --- a pre-processing step --- in automated theorem provers, which also involve equivalences that are only classically valid.
B) We will identify assumptions under which our adaptions of the techniques 1) and 2) guarantee that such a program extraction is possible.
We are not aware of related results in the literature.
